Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-06-14 MS/ms.Rmd
More writing of applesa nd oranges problem

Meta-analysis is a well-developed research field with a long history in statistics, medicine, psychology and ecology [@Gurevitch2018; @Nakagawa2017]. It has emerged as the 'gold-standard' across various disciplines for quantitative synthesis and is becoming increasingly more prominant the field of comparative physiology. Formal meta-analyses involve a systematic literature-based search, extraction of data from studies meeting specific inclusion criteria, and the subsequent calculation of effects sizes from summary statistics provided by papers. The entire process should follow and report upon this entire process carefully to ensure the meta-analysis is as transparent and reproducible as possible [i.e., see the Preferred Reporting Items for Systematic Reviews and Meta-Analyses in Ecology and Evolution (PRISMAEcoEvo) protocol, @ODea2021]. Aside from providing an up-to-date overview of the research question, along with identifying any gaps and biases, there are two major goals of meta-analysis: 1) to provide an overall mean estimate of a treatment effect or relationship and 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013]. A critically important feature of meta-analaysis is its ability to account for sampling error [@Koricheva2013]. The inverse sampling error is used to weight higher powered studies in the analysis and make overall estimates more precise. While many meta-analyses might employ unweighted analyses, these studies will not gain greater precision on estimates because sampling variance cannot be formally accounted for in the analysis, even if the overall mean itself is unbiased.

@@

## **The Apples and Oranges 'Problem': How lack of comparability can make effects vary widely across studies**
 I'm thinking now, in re-reading this, that we could use a slightly expanded version of paragraph 4 taht describes this problem in a little more detail. This is likely the place for a figure, which I am having some trouble envisaging where one is best placed. It just otherwise seems like a bit of a jump to otherwise head straight into the survey. Here we should also talk about quality control, the need to be transparent and have careful reporting. This is a big issue in medicine. Talked about a lot. Integrate blurb below...try to take a balanced approach.

In the words of Gene Glass, arguably the father of meta-analysis: "Of course it mixes apples and oranges; in the study of fruit nothing else is sensible; comparing apples and oranges is the only endeavor worthy of true scientists; comparing apples to apples is trivial" [p.g., 224, @Glass2015]. @Glass2015's intention was to simply highlight that whether comparing apples and oranges is a good idea simply boils down to the specific question of interest. If we are interested in gaining insight into 'fruit' then synthesising apples and oranges is a good idea. Instead, if we are intersted in generalising to apples only, then mixing fruit will not be ideal. We tend to agree with @Glass2015 and his sentiment is of particular relevance to comparative physiology. By synthesizing diverse study outcomes, it allows us to not just report on overall effects, but understand critical features of the literature, study systems and approaches that can explain the diversity of effect outcomes observed. In doing so, it provides a richer set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included within a meta-analysis. Answers to certain questions will inevitably depend on the homogeneity of effects being synthesized.

>>>>

Meta-analysis is a well-developed research field with a long history in statistics, medicine, psychology and ecology [@Gurevitch2018; @Nakagawa2017; @Glass2015]. It has emerged as the 'gold-standard' across various disciplines for quantitative synthesis and is becoming increasingly more prominant the field of comparative physiology. Formal meta-analyses involve a systematic literature-based search, extraction of data from studies meeting specific inclusion criteria, and the subsequent calculation of effects sizes from summary statistics provided by papers. The entire process should follow and report upon this entire process carefully to ensure the meta-analysis is as transparent and reproducible as possible [i.e., see the Preferred Reporting Items for Systematic Reviews and Meta-Analyses in Ecology and Evolution (PRISMAEcoEvo) protocol, @ODea2021]. Aside from providing an up-to-date overview of the research question, along with identifying any gaps and biases, there are two major goals of meta-analysis: 1) to provide an overall mean estimate of a treatment effect or relationship and 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013]. A critically important feature of meta-analaysis is its ability to account for sampling error [@Koricheva2013]. The inverse sampling error is used to weight higher powered studies in the analysis and make overall estimates more precise. While many meta-analyses might employ unweighted analyses, these studies will not gain greater precision on estimates because sampling variance cannot be formally accounted for in the analysis, even if the overall mean itself is unbiased.

@@

## **The Apples and Oranges 'Problem': How lack of comparability can make effects vary widely across studies**
 I'm thinking now, in re-reading this, that we could use a slightly expanded version of paragraph 4 taht describes this problem in a little more detail. This is likely the place for a figure, which I am having some trouble envisaging where one is best placed. It just otherwise seems like a bit of a jump to otherwise head straight into the survey. Here we should also talk about quality control, the need to be transparent and have careful reporting. This is a big issue in medicine. Talked about a lot. Integrate blurb below...try to take a balanced approach. *I will add an introduction bit to this paragraph below to put into context....*

In the words of Gene Glass, the father of meta-analysis: "Of course it mixes apples and oranges; in the study of fruit nothing else is sensible; comparing apples and oranges is the only endeavor worthy of true scientists; comparing apples to apples is trivial" [p.g., 224, @Glass2015]. @Glass2015's intention was to simply highlight that whether comparing apples and oranges is a good idea simply boils down to the specific question of interest. If we are interested in gaining insight into 'fruit' then synthesising apples and oranges is the only thing that makes sense. Instead, if we are interested in generalising only to apples then mixing fruit will not be ideal. We tend to agree with @Glass2015, and his sentiment is of particular relevance to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms then including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are known to differ. By synthesizing diverse study outcomes, it allows us to not just report on overall effects (if the question is relevant to ask), but understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction) and approaches that can explain the diversity of effect outcomes observed. In doing so, it provides a richer set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included within a meta-analysis - an important reason why transparency and reproducibility are such a prominant feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a polluntant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant, and indeed may be misleading. It is important to consider these questions when conducting meta-analyses.

Of course, there are certain types of heterogenity that transcend different questions. In other words, if we compare different types of 'apples' is it still fair to synthesize these studies to understand overall effects? This situation might be analogous to a meta-analysis in comparative physiology that attemps to understand the effects of temperature on fish growth. Studies might rear populations of fish under different temperatures, but in the end, they are still all fish, governed by the same thermodynamic effects on growth. In this situation, we might still be interested in understanding overall effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. However, such a study does not preclude us from also exploring important moderators or predictors of variability in effects across studies.
